Ultrafast Control of Spin-Spin Interactions in Two-Dimensional Magnetic Systems: Kane-Mele-Hubbard Model

### 2019-03-05 Paper/paper.tex
first version 12:00-05March

\begin{abstract}
We study ultrafast optical control of exchange and Dzyaloshinskii-Moriya interactions in two-dimensional honeycomb lattices described by the Kane-Mele-Hubbard model at half-filling and strongly correlated limit, i.e., Mott insulator regime of a canted antiferromagnet. Based on Floquet theory calculations, we show that by changing the amplitude and frequency of laser pulses, it is possible to tune the amplitude, the sign and even the ratio between these two spin interactions. Our results pave the way for ultrafast optical manipulation of spins in recently discovered two-dimensional magnetic materials.
\end{abstract}

@@

\textit{Introduction}.--- Discovery of athermal all-optical control of antiferromagnetic (AFM) and ferromagnetic (FM) spins by intense and high-frequency laser pulses has elevated spintronics to a new era called ultrafast magnetism. Although the underlying microscopic mechanism behind this phenomena is unclear yet, there have been many attempts to uncover its origins. Recently, it has been theoretically proposed that laser pulses can modify and even reverse the sign of exchange interaction, the strongest spin interaction in magnetic systems. This finding has been indirectly supported by a few recent experiments. Another important spin interaction is Dzyaloshinskii-Moriya intercation (DMI), an antisymmetric exchange interaction appeared in inversion symmetry broken systems, that breaks the chiral symmetry of the system. While this interaction is much weaker than the exchange interaction, its presence has important consequences in magnetic materials such as weak ferromagnetism in AFM materials, stabilizing topological objects like chiral skyrmions and domain walls, and exotic phase of topological magnon insulators. The ratio between these two interactions control the tilting angle of canted spins. Proposing a mechanism to tune this ratio has important consequences in ultrafast spin dynamics and switching. In this Letter, we show that intense and high-frequency laser pulses can change this ratio in a class of two-dimensional (2D) magnetic materials described by the Kane-Mele-Hubbard model.

2D van der Waals magnetic materials have been recently discovered with metallic, semiconducting and insulating band structures in both AFM and FM phases. We can now use the advantage of both low dimensionality and magnetic order in the same material for the first time which promises novel spintronics devices with exceptional performances.

\textit{Model Hamiltonian}.---
Effective Hamiltonian of a 2D planar honeycomb lattice in the presence of electron-electron interactions can be described by Kane-Mele-Hubbard model \cite{Kane2005,Griset2012,Auerbach}. 
The Hamitlonian in the absence of external perturbations is the sum of the kinetic term $\hat{\mathcal{H}}_t$, the intrinsic SOI $\hat{\mathcal{H}}_{\text{SOI}}$ and electron-electron interaction modelled by extended Hubbard model $\hat{\mathcal{H}}_{\text{int}}$,
\begin{equation}
\label{MKMH}
\hat{H}_0 = \hat{\mathcal{H}}_{\text{K}} + \hat{\mathcal{H}}_{\text{SOI}} + \hat{\mathcal{H}}_{\text{int}},
\end{equation}
where
\begin{align}
\hat{\mathcal{H}}_{\text{K}} &= - \sum_{\langle i,j \rangle, \tau} t_{1}\hat{c}_{i \tau}^\dagger \hat{c}_{j \tau} - \sum_{\langle \langle i,j \rangle \rangle, \tau} t_{2}\hat{c}_{i \tau}^\dagger \hat{c}_{j \tau} \label{Ht} \\
\hat{\mathcal{H}}_{\text{SOI}} & = \sum_{\langle \langle i,j \rangle \rangle, \tau,\tau'} i\Delta\nu_{ij}\sigma^z_{\tau, \tau'}\hat{c}_{i \tau}^\dagger \hat{c}_{j \tau'} \label{Hsoi} \\
\hat{\mathcal{H}}_{\text{int}} &= U_{00}\sum_{i=1} \hat{n}_{i\uparrow}\hat{n}_{i\downarrow} + \frac{1}{2}\sum_{\langle i,j \rangle, \tau \tau'} V_{ij}\hat{n}_{i\tau}\hat{n}_{j\tau'} \label{Hint}.
\end{align}
where $\langle\cdot\rangle$ and $\langle\langle\cdot\rangle\rangle$ denote, respectively, nearest neighbors (NN) and next-nearest neighbours (NNN), $\hat{c}_{i \sigma}^\dagger$ ($ \hat{c}_{i \sigma}$) are fermionic creation (annihilation) operators for an electron on site $i$ and spin state $\tau=\{\uparrow,\downarrow\}$, $t_1$ and $t_2$ are NN and NNN hopping amplitudes, $\Delta$ is the intrinsic SOI constant, $\nu_{ij}=\pm 1$ depending on whether the electron traversing from $i$ to $j$ makes a right ($+1$) or a left turn ($-1$), $\sigma^{z}$ is the $z$-component of Pauli matrices $\bm{\sigma}$, $U_{00}$ and $V_{ij}$ are the on-site and NN Coulomb interactions.
The presence of the intrinsic SOI, Eq. (\ref{Hsoi}), reduces the $\mathrm{SU}(2)$ symmetry of a pure Hubbard model to the $\mathrm{U}(1)$ spin group.

@@

Using the variational principle, it has been shown that the NN Coulomb interaction can be effectively approximated by a renormalized local interaction $U =U_{00} - \bar{V}$, where $\bar{V}$ is a weighted average of the NN Coulomb interaction \cite{Schuler2013,Stepanov2017}. Thus, we only take into account the local Coulomb interaction in our total Hamiltonian and rewrite the interaction part of the Hamiltonian (\ref{Hint}),
\begin{equation}
\hat{\mathcal{H}}_{\text{int}} \approx U\hat{d},
\end{equation}
where we define the doublon number operator $\hat{d} = \sum_{i=1} \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}$ with eigenvalues $d$ and projection operators $\hat{P}_d$. We are interested to Kane-Mele-Hubbard model at half-filling in strong-coupling regime $U\gg t_{1(2)}$. In the strong-coupling limit any state with a nonzero number of double occupancies ($d \neq 0$) will have much larger energy than those with $d=0$. We obtain an effective Hamiltonian acting on the $d=0$ subspace by standard second-order perturbation theory in the hopping terms. Using the following relations,
\begin{align}
\hat{c}_{i \tau}^\dagger \hat{c}_{i \tau'} &= \frac{1}{2} (n_{i \uparrow} + n_{i \downarrow})\delta_{\tau \tau'}  + \bs{S}_i\cdot\bs{\tau}_{\tau', \tau}, \label{SpinOperatorInv1}\\
\hat{c}_{i \tau} \hat{c}_{i \tau'}^\dagger &= \frac{1}{2} (2 - n_{i \uparrow} - n_{i \downarrow}) \delta_{\tau \tau'} - \bs{S}_i\cdot\bs{\tau}_{\tau, \tau'}, \label{SpinOperatorInv2}
\end{align}
we find the spin Hamiltonian for 2D AFM Mott insulators,
\begin{align}
\label{MKMHeff0}
H_{\text{S}} =& J_{1}\sum_{\langle i,j \rangle} \bs{S}_i\cdot\bs{S}_j + J_{2}\sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i\cdot\bs{S}_j\n \\
&+ \sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i \bs{\Gamma}_{ij} \bs{S}_j +\sum_{\langle \langle i,j \rangle \rangle} \bs{D}_{ij}\cdot \bs{S}_i \times \bs{S}_j,
\end{align}
with the following spin-spin interactions,
\begin{align}
J_{1(2)} &= \frac{2t_{1(2)}^2}{U}, \\
\bs{\Gamma}_{ij} &=\frac{2\Delta^2}{U} \text{diag}(-1,-1,1),\\
\bs{D}_{ij} &= - \frac{4 t_2 \Delta}{U}\nu_{ij}  \hat{\mathrm{e}}_z.
\end{align}
The first and second terms in the spin Hamiltonian, Eq. (\ref{MKMHeff0}), are the NN and NNN symmetric Heisenberg AFM exchange interactions, $J_{1(2)}>0$, respectively; the third term is a NNN Heisenberg XXZ-like term arising from the intrinsic SOI, and finally, the last term is the intrinsic NNN DMI. The intrinsic SOI in Eq. (\ref{Hsoi}), leads to a DMI vector, $\bs{D}_{ij}$, perpendicular to the honeycomb plane while it is straightforward to show that the Rashba SOI, arising from breaking the inversion symmetry of the honeycomb lattice, results an inplane DMI vector perpendicular to the lattice bonds.
For the sake of completeness, we explain shortly the effect of disorders by adding an on-site disorder $\sum_{i \tau} \varepsilon_i \hat{c}_{i \tau}^\dagger \hat{c}_{i \tau}$ in the Kane-Mele-Hubbard Hamiltonian \ref{MKMH}, where $\varepsilon_i$ is an uncorrelated random variable. Following the above procedure, it can be shown that the spin interaction parameters are only renormalized by $U^{-1} \rightarrow U/\left(U^2-(\epsilon_j-\epsilon_i)^2\right)$ \cite{Protopopov2018}. In the large $U$ limit, the effect of disorder is negligible and we thus ignore it in the rest of this Letter.

@@

There are some recent experiments on possibility of ultrafast optical modification of exchange interactions in the bulk of iron oxides \cite{Mikhaylovskiy2015,Kim2016,Jin2014}. We hope that our work stimulate new experiments on measuring exchange interaction and DMI in 2D magnetic systems.

>>>>

\begin{abstract}
We study ultrafast optical control of exchange and Dzyaloshinskii-Moriya interactions in two-dimensional honeycomb lattices described by the Kane-Mele-Hubbard model at half-filling and strongly correlated limit, i.e., Mott insulator regime of a canted antiferromagnet. Based on Floquet theory calculations, we show that by changing the amplitude and frequency of polarized laser pulses, it is possible to tune the amplitude, the sign and even the ratio between these two spin interactions. We find that in this model, the renormaization of spin interactions are helicity independent. Our results pave the way for ultrafast optical manipulation of spins in recently discovered two-dimensional magnetic materials.
\end{abstract}

@@

\textit{Introduction}.--- Discovery of athermal all-optical control of order parameters in antiferromagnetic (AFM) and ferromagnetic (FM) materials by intense and high-frequency laser pulses has elevated spintronics to a new era called ultrafast magnetism. Although the underlying microscopic mechanism behind this phenomena is unclear yet, there have been many attempts to uncover its origins. Recently, it has been theoretically proposed that laser pulses can modify and even reverse the sign of exchange interaction, the strongest spin interaction in magnetic systems. This finding has been indirectly supported by a few recent experiments. Another important spin interaction is Dzyaloshinskii-Moriya intercation (DMI), an antisymmetric exchange interaction appeared in inversion symmetry broken systems, that breaks the chiral symmetry of the system. While this interaction is much weaker than the exchange interaction, its presence has important consequences in magnetic materials such as weak ferromagnetism in AFM materials, stabilizing topological objects like chiral skyrmions and domain walls, and exotic phase of topological magnon insulators. The ratio between these two interactions control the tilting angle of canted spins. Proposing a mechanism to tune this ratio has important consequences in ultrafast spin dynamics and switching. In this Letter, we show that intense and high-frequency laser pulses can change this ratio in a class of two-dimensional (2D) magnetic materials described by the Kane-Mele-Hubbard model.

2D van der Waals magnetic materials with metallic, semiconducting and insulating band structures have been recently discovered in both AFM and FM phases. We can now use the advantage of both low dimensionality and magnetic order in the same material for the first time which promises novel spintronics devices with exceptional performances.

\textit{Model Hamiltonian}.---
Effective Hamiltonian of a 2D planar honeycomb lattice in the presence of electron-electron interactions can be described by Kane-Mele-Hubbard model \cite{Kane2005,Griset2012,Auerbach}. 
The Hamitlonian in the absence of external perturbations is the sum of the kinetic term $\hat{\mathcal{H}}_t$, the intrinsic SOI $\hat{\mathcal{H}}_{\text{SOI}}$ and electron-electron interaction modelled by extended Hubbard model $\hat{\mathcal{H}}_{\text{int}}$,
\begin{equation}
\label{MKMH}
\hat{H}_0 = \hat{\mathcal{H}}_{\text{K}} + \hat{\mathcal{H}}_{\text{SOI}} + \hat{\mathcal{H}}_{\text{int}},
\end{equation}
where
\begin{align}
\hat{\mathcal{H}}_{\text{K}} &= - \sum_{\langle i,j \rangle, \tau} t_{1}\hat{c}_{i \tau}^\dagger \hat{c}_{j \tau} - \sum_{\langle \langle i,j \rangle \rangle, \tau} t_{2}\hat{c}_{i \tau}^\dagger \hat{c}_{j \tau} \label{Ht} \\
\hat{\mathcal{H}}_{\text{SOI}} & = \sum_{\langle \langle i,j \rangle \rangle, \tau,\tau'} i\Delta\nu_{ij}\sigma^z_{\tau, \tau'}\hat{c}_{i \tau}^\dagger \hat{c}_{j \tau'} \label{Hsoi} \\
\hat{\mathcal{H}}_{\text{int}} &= U_{00}\sum_{i=1} \hat{n}_{i\uparrow}\hat{n}_{i\downarrow} + \frac{1}{2}\sum_{\langle i,j \rangle, \tau \tau'} V_{ij}\hat{n}_{i\tau}\hat{n}_{j\tau'} \label{Hint}.
\end{align}
where $\langle\cdot\rangle$ and $\langle\langle\cdot\rangle\rangle$ denote, respectively, nearest neighbors (NN) and next-nearest neighbours (NNN), $\hat{c}_{i \sigma}^\dagger$ ($ \hat{c}_{i \sigma}$) are fermionic creation (annihilation) operators for an electron on site $i$ and spin state $\tau=\{\uparrow,\downarrow\}$, $t_1$ and $t_2$ are NN and NNN hopping amplitudes, $\Delta$ is the intrinsic SOI constant, $\nu_{ij}=\pm 1$ depending on whether the electron traversing from $i$ to $j$ makes a right ($+1$) or a left turn ($-1$), $\sigma^{z}$ is the $z$-component of Pauli matrices $\bm{\sigma}$, $U_{00}$ and $V_{ij}$ are the on-site and NN Coulomb interactions.
The presence of the intrinsic NNN SOI, Eq. (\ref{Hsoi}), reduces the $\mathrm{SU}(2)$ symmetry of original Hubbard model to the $\mathrm{U}(1)$ spin group. In buckled noncoplanar honeycomb lattices and systems with structural inversion asymmetry, presence of, respectively, NNN and NN Rashba SOIs, furthur reduces the symmetry to $Z_2$.

@@

Using the variational principle, it has been shown that the NN Coulomb interaction can be effectively approximated by a renormalized local interaction $U =U_{00} - \bar{V}$, where $\bar{V}$ is a weighted average of the NN Coulomb interaction \cite{Schuler2013,Stepanov2017}. Thus, we only take into account the local Coulomb interaction in our total Hamiltonian and rewrite the interaction part of the Hamiltonian (\ref{Hint}),
\begin{equation}
\hat{\mathcal{H}}_{\text{int}} \approx U\hat{d},
\end{equation}
where we define the doublon number operator $\hat{d} = \sum_{i=1} \hat{n}_{i\uparrow}\hat{n}_{i\downarrow}$ with eigenvalues $d$ and projection operators $\hat{P}_d$. We are interested to Kane-Mele-Hubbard model at half-filling in strongly correlated regime $U\gg t_{1(2)}$. In such strong-coupling limit any state with a nonzero number of double occupancies ($d \neq 0$) will have much larger energy than those with $d=0$. We obtain an effective Hamiltonian acting on the $d=0$ subspace by standard second-order perturbation theory in the hopping terms. Using the following relations,
\begin{align}
\hat{c}_{i \tau}^\dagger \hat{c}_{i \tau'} &= \frac{1}{2} (n_{i \uparrow} + n_{i \downarrow})\delta_{\tau \tau'}  + \bs{S}_i\cdot\bs{\tau}_{\tau', \tau}, \label{SpinOperatorInv1}\\
\hat{c}_{i \tau} \hat{c}_{i \tau'}^\dagger &= \frac{1}{2} (2 - n_{i \uparrow} - n_{i \downarrow}) \delta_{\tau \tau'} - \bs{S}_i\cdot\bs{\tau}_{\tau, \tau'}, \label{SpinOperatorInv2}
\end{align}
we find the spin Hamiltonian for 2D AFM Mott insulators,
\begin{align}
\label{MKMHeff0}
H_{\text{S}} =& J_{1}\sum_{\langle i,j \rangle} \bs{S}_i\cdot\bs{S}_j + J_{2}\sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i\cdot\bs{S}_j\n \\
&+ \sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i \bs{\Gamma}_{ij} \bs{S}_j +\sum_{\langle \langle i,j \rangle \rangle} \bs{D}_{ij}\cdot \bs{S}_i \times \bs{S}_j,
\end{align}
with the following spin-spin interactions,
\begin{align}
J_{1(2)} &= \frac{2t_{1(2)}^2}{U}, \\
\bs{\Gamma}_{ij} &=\frac{2\Delta^2}{U} \text{diag}(-1,-1,1),\\
\bs{D}_{ij} &= - \frac{4 t_2 \Delta}{U}\nu_{ij}  \hat{\mathrm{e}}_z.
\end{align}
The first and second terms in the spin Hamiltonian, Eq. (\ref{MKMHeff0}), are the NN and NNN symmetric Heisenberg AFM exchange interactions, $J_{1(2)}>0$, respectively; the third term is a NNN Heisenberg XXZ-like term arising from the intrinsic SOI, and finally, the last term is the intrinsic NNN DMI. The intrinsic SOI in Eq. (\ref{Hsoi}), leads to a DMI vector, $\bs{D}_{ij}$, perpendicular to the honeycomb plane while it is straightforward to show that the Rashba SOI results an inplane DMI vector perpendicular to the lattice bonds.
For the sake of completeness, we explain shortly the effect of disorders by adding an on-site disorder $\sum_{i \tau} \varepsilon_i \hat{c}_{i \tau}^\dagger \hat{c}_{i \tau}$ in the Kane-Mele-Hubbard Hamiltonian \ref{MKMH}, where $\varepsilon_i$ is an uncorrelated random variable. Following the above procedure, it can be shown that the spin interaction parameters are only renormalized by $U^{-1} \rightarrow U/\left(U^2-(\epsilon_j-\epsilon_i)^2\right)$ \cite{Protopopov2018}. In the large $U$ limit, the effect of disorder is negligible and we thus ignore it in the rest of this Letter.

@@

There are some recent experiments on possibility of ultrafast optical modification of exchange interactions in the bulk of iron oxides \cite{Mikhaylovskiy2015,Kim2016,Jin2014}. We hope that our work stimulate new experiments on measuring both exchange interaction and DMI in 2D magnetic systems.

In summary, using Floquet theory we have investigated the effect of polarized and intense high-frequency laser pulses on 2D canted antiferromagnetic Mott insulators. We have found that both amplitude and sign of the ratio between the exchange interaction and DMI is changed in honeycomb lattices irrespective to the helicity of laser pulses. Our calculations open a new way to ultrafast energy-efficient control of spins and engineering of topological objects and topological properties of 2D van der Waals magnetic materials.
